Analysis of storm damages across the US
========================================================


## Synopsis

For this project (as part of the "Data Science" course offered by Coursera), data from the  U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from 1950 to November 2011 has been analysed to find the worst event types regarding death toll and economic damage. The event type with the highest number of fatalities has been found to be tornados, while floods have caused the highest economic damage. 


## Data Processing

First, we read the data into R:

```{r, cache=TRUE}
rawdata <- read.csv(file="repdata_data_StormData.csv", header=TRUE, skip=0);
```

Sum the fatalities and injuries for all event types:

```{r, cache=TRUE}
fatalities <- aggregate(FATALITIES ~ EVTYPE, FUN=sum, data=rawdata);
injuries <- aggregate(INJURIES ~ EVTYPE, FUN=sum, data=rawdata);
```

Define a function to convert PROPDMGEXP and CROPDMGEXP to numeric values

```{r}
exp2num <- function(exponent)
{
    ifelse(exponent=="K", 1000, ifelse(exponent=="M", 1000000, ifelse(exponent=="B", 1000000000, 1)));
}
```

Sum the property damage and crop damage for all event types:

```{r, cache=TRUE}
rawdata$PROPERTYDMG <- rawdata$PROPDMG*exp2num(rawdata$PROPDMGEXP);
rawdata$CROPDAMAGE <- rawdata$CROPDMG*exp2num(rawdata$CROPDMGEXP);
propertydamage <- aggregate(PROPERTYDMG ~ EVTYPE, FUN=sum, data=rawdata);
cropdamage <- aggregate(CROPDAMAGE ~ EVTYPE, FUN=sum, data=rawdata);
```

Calculate the total damage (property plus crop) for all event types:

```{r}
totaldamage <- propertydamage;
totaldamage$PROPERTYDMG <- NULL;
totaldamage$DAMAGE <- propertydamage$PROPERTYDMG + cropdamage$CROPDAMAGE;
```

Extract the event types with nonzero fatalities / injuries / total damage and order them in decreasing order:

```{r}
fatal <- fatalities[fatalities$FATALITIES>0,];
fatal <- fatal[order(fatal$FATALITIES, decreasing=TRUE),];
inj <- injuries[injuries$INJURIES>0,];
inj <- inj[order(inj$INJURIES, decreasing=TRUE),];
damage <- totaldamage[totaldamage$DAMAGE>0,];
damage <- damage[order(damage$DAMAGE, decreasing=TRUE),];
```


## Results

The event type with the greatest economic consequences has been found to be `r damage$EVTYPE[1]` with a total damage (including both property and crops damage) worth `r damage$DAMAGE[1]/1e+9` billion USD. [inline code: `damage$EVTYPE[1]` and `damage$DAMAGE[1]/1e+9` ]. A plot of the "top ten" event types by total damage can be found in the following figure.

```{r fig.width=7, fig.height=6, cache=FALSE}
par(mar=c(8,4.1,4.1,2.1), cex.axis=0.7);
plot(damage$DAMAGE[1:10]/1e+9, xaxt="n", yaxt="n", xlab="", ylab="total damage (in billion USD)", pch=16, main="total damage per \"top ten\" event types");
axis(1, at=1:10, labels=damage$EVTYPE[1:10], las=2);
axis(2, las=1);
```
 
From the data, the most harmful event type regarding public health (where injuries are ignored, i.e. fatalities are weighted infinitely more strongly than injuries) is `r fatal$EVTYPE[1]` with a total death toll of `r fatal$FATALITIES[1]`. [inline code: `fatal$EVTYPE[1]` and `fatal$FATALITIES[1]` ]. A plot of the "top ten" event types by death toll is shown in the following figure:

```{r fig.width=7, fig.height=6, cache=FALSE}
par(mar=c(8,4.1,4.1,2.1), cex.axis=0.7);
plot(fatal$FATALITIES[1:10]/1e+3, xaxt="n", yaxt="n", xlab="", ylab="fatalities (in thousands)", pch=16, main="fatalities per \"top ten\" event types");
axis(side=1, at=1:10, labels=fatal$EVTYPE[1:10], las=2);
axis(2, las=1);
```

