Analysis of storm damages across the US
========================================================


## Synopsis


## Data Processing

First, we read the data into R:

```{r, cache=TRUE}
rawdata <- read.csv(file="repdata_data_StormData.csv", header=TRUE, skip=0);
```

Sum the fatalities and injuries for all event types:

```{r, cache=TRUE}
fatalities <- aggregate(FATALITIES ~ EVTYPE, FUN=sum, data=rawdata);
injuries <- aggregate(INJURIES ~ EVTYPE, FUN=sum, data=rawdata);
```

Define a function to convert PROPDMGEXP and CROPDMGEXP to numeric values

```{r}
exp2num <- function(exponent)
{
    ifelse(exponent=="K", 1000, ifelse(exponent=="M", 1000000, ifelse(exponent=="B", 1000000000, 1)));
}
```

Sum the property damage and crop damage for all event types:

```{r, cache=TRUE}
rawdata$PROPERTYDMG <- rawdata$PROPDMG*exp2num(rawdata$PROPDMGEXP);
rawdata$CROPDAMAGE <- rawdata$CROPDMG*exp2num(rawdata$CROPDMGEXP);
propertydamage <- aggregate(PROPERTYDMG ~ EVTYPE, FUN=sum, data=rawdata);
cropdamage <- aggregate(CROPDAMAGE ~ EVTYPE, FUN=sum, data=rawdata);
```

Calculate the total damage (property plus crop) for all event types:

```{r}
totaldamage <- propertydamage;
totaldamage$PROPERTYDMG <- NULL;
totaldamage$DAMAGE <- propertydamage$PROPERTYDMG + cropdamage$CROPDAMAGE;
```

Extract the event types with nonzero fatalities / injuries / total damage and order them in decreasing order:

```{r}
fatal <- fatalities[fatalities$FATALITIES>0,];
fatal <- fatal[order(fatal$FATALITIES, decreasing=TRUE),];
inj <- injuries[injuries$INJURIES>0,];
inj <- inj[order(inj$INJURIES, decreasing=TRUE),];
damage <- totaldamage[totaldamage$DAMAGE>0,];
damage <- damage[order(damage$DAMAGE, decreasing=TRUE),];
```


## Results

```{r fig.width=7, fig.height=6}
par(mar=c(8,4.1,4.1,2.1), cex.axis=0.7);
plot(damage$DAMAGE[1:10]/1e+9, xaxt="n", xlab="", ylab="total damage (in billion USD)", pch=16, main="total damage per event type");
axis(1, at=1:10, labels=damage$EVTYPE[1:10], las=2);

```



```{r fig.width=7, fig.height=6}
par(mar=c(8,4.1,4.1,2.1), cex.axis=0.7);
plot(fatal$FATALITIES[1:10]/1e+3, xaxt="n", xlab="", ylab="fatalities (in thousands)", pch=16, main="total fatalities per event type");
axis(1, at=1:10, labels=damage$EVTYPE[1:10], las=2);

```




```{r}

```


```{r fig.width=7, fig.height=6}

```

